Carrier phase and amplitude estimation of $M$-ary phase shift keying using pilots and data

### 2012-05-12 paper.tex
further edits to paper.  Just realised some simplifications too!

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, zero mean, circularly symetric complex random variables and let $\{y_i, i \in P \cup D\}$ be observed noisy symbols given by the signal model~\eqref{eq:sigmod}.   Let $\hat{a}$ be the least squares estimator of $a_0$ and put $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number.  Put $L = \abs{P \cup D}$ and define real numbers $p$ and $d$ as the limits
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [-\pi,\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{1}{\rho_0} w_i e^{-j\theta_0} s_i^*,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x)
\]
where
\begin{align*}
h_1(x) &= \prob R_1 \cos(x + \Phi_1) \;\;\; \text{and}, \\
h_2(x) &=  \prob R_1 \cos\sfracpart{ x + \Phi_1}.
\end{align*}
If $p > 0$ and if $\abs{G(x)}$ is uniquely maximised at $x = 0$ over the interval $[-\pi,-\pi)$, then:
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

\begin{IEEEproof} (Theorem~\ref{thm:normality})
The proof is complicated by the fact that $G_L$ is not differentiable everywhere due to the function $\fracpart{\cdot}$ not being differentiable at half integers.  This precludes the use of ``standard approaches'' to proving normality that are based on the mean value theorem~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats,Pollard_new_ways_clts_1986,Pollard_conv_stat_proc_1984,Pollard_asymp_empi_proc_1989}.  However, we show in Lemma~\ref{lem:diffatlambdaL} that the derivative does exist, and is equal to zero, at the maximiser $\hat{\lambda}_L$, so
\begin{align*}
G_L^\prime(\hat{\lambda}_L) &= 0 \\
&= \frac{1}{L} \sum_{i \in P} R_i \sin(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \sin\sfracpart{\lambda + \Phi_i}.
\end{align*}
Define the function $Q(\lambda) = \prob G_L^\prime(\lambda)$ and write
\begin{align*}
0 &= G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) + Q(\hat{\lambda}_L) \\
&= \sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) + \sqrt{L}Q(\hat{\lambda}_L).
\end{align*}
Lemma~\ref{lem:Qconv} shows that
\[
\sqrt{L} Q(\hat{\lambda}_L) = \sqrt{L} Q(0) + \sqrt{L} \hat{\lambda}_L\big( ?  + o_P(1) \big)
\]
where $o_P(1)$ denotes a number converging in probablity to zero as $L \rightarrow \infty$.  Lemma~\ref{lem:empiricprocc} shows that
\[
\sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} \big( G_L^\prime(0) - Q(0) \big).
\]
It follows that
\[
0 = o_P(1) + \sqrt{L}G_L^\prime(0) + \sqrt{L} \hat{\lambda}_L \big( ?  + o_P(1) \big)
\]
and rearranging gives,
\[
\sqrt{L} \hat{\lambda}_L = o_P(1) + \frac{\sqrt{L}G_L'(0)}{?  + o_P(1)}.
\]
Lemma~\ref{lem:convdistGLdash} shows that the distribution of $\sqrt{L}G_L'(0)$ converges to the normal with zero mean and variance
\[
p \prob R_1^2 \sin^2(\Phi_1) + d \prob R_1^2 \sin^2\fracpart{\Phi_1}
\]
and it follows that the distribution of $\sqrt{L}\hat{\lambda}_L$ converges to the normal with zero mean and variance
\[
\frac{p \prob R_1^2 \sin^2(\Phi_1) + d \prob R_1^2 \sin^2\fracpart{\Phi_1}}{?}
\]
as required.

@@

\begin{lemma}\label{lem:Qconv}
\[
\sqrt{L} Q(\hat{\lambda}_L) = \sqrt{L} Q(0) + \sqrt{L} \hat{\lambda}_L\big( ?  + o_P(1) \big)
\]
\end{lemma}
\begin{IEEEproof}

\end{IEEEproof}

\begin{lemma}\label{lem:empiricprocc} BLERG
\[
\sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} \big( G_L^\prime(0) - Q(0) \big).
\]
\end{lemma}
\begin{IEEEproof}
\end{IEEEproof}

\begin{lemma}\label{lem:convdistGLdash}
The distribution of $\sqrt{L}G_L'(0)$ converges to the normal with zero mean and variance
\[
p \prob R_1^2 \sin^2(\Phi_1) + d \prob R_1^2 \sin^2\fracpart{\Phi_1}.
\]
\end{lemma}
\begin{proof}
Observe that 
\[
\sqrt{L} G_L^\prime(0) = \frac{1}{\sqrt{L}} \sum_{i \in P} R_i \sin(\Phi_i) + \frac{1}{\sqrt{L}} \sum_{i \in D} R_i \sin\sfracpart{\Phi_i}.
\]
It follows from the standard central limit that the distribution of
\[
\frac{1}{\sqrt{L}} \sum_{i \in P} R_i \sin(\lambda + \Phi_i) = \sqrt{\frac{\abs{P}}{L}} \frac{1}{\sqrt{\abs{P}}} \sum_{i \in P} R_i \sin(\Phi_i) 
\]
converges to the normal with mean $k_1(0) = 0$ and variance 
\[
p \prob R_1^2 \sin^2(\Phi_1).
\]
Similarly, the distribution of 
\[
\frac{1}{\sqrt{L}} \sum_{i \in D} R_i \sin(\lambda + \Phi_i) = \sqrt{\frac{\abs{D}}{L}} \frac{1}{\sqrt{\abs{D}}} \sum_{i \in D} R_i \sin\fracpart{\Phi_i} 
\]
converges to the normal with mean $k_2(0) = 0$ and variance
\[
d \prob R_1^2 \sin^2\fracpart{\Phi_1}.
\]
\end{proof}

@@

% \section{A linear-time algorithm}\label{sec:line-time-algor}

% \begin{lemma}
% Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ so that $\hat{\theta}$ is the least squares estimator of the carrier phase and $\hat{\rho}$ is the least squares estimator of the amplitude.  Let 
% \[
% \lambda_i = \phi_i - \hat{\theta} - \frac{2\pi}{M}\round{\frac{M}{2\pi}(\phi_i - \hat{\theta})}.
% \]
% Then 
% \[
% \abs{\lambda_i} \leq \frac{1}{2M} - \frac{1}{2LM BLERG}
% \]
% for all $i \in P \cup D$.
% \end{lemma}
% \begin{proof}
% Recall that $\hat{d}_i = \hat{d}_i(\hat{\theta})$ are the minimsers of $SS$.  The proof now proceeds by contradiction.  Assume that $\lambda_i > \frac{1}{2M} - \frac{1}{2LM BLERG}$ for some $k$.  Put $r_i = \hat{d}_i e_k$, then
% \begin{align*}
% Y(\{ r_i, i \in D\}) &= \frac{1}{L}\left( B + \sum_{i \in D} y_i\hat{r}_i^* \right) \\
% &= \hat{Y} + \frac{1}{L}\eta y_k\hat{d}_k^*.
% \end{align*}
% where we put $\hat{Y} = Y(\{ \hat{d}_i, i \in D\})$.  Now,
% \begin{align*}
% SS(\{r_i, i \in D\}) &= A - \frac{1}{L} \abs{ Y(\{ r_i, i \in D\}) }^2 \\
% &= A - \frac{1}{L} \abs{ \hat{Y} + \frac{1}{L} \eta y_k\hat{d}_k^* }^2 \\
% &= A - \frac{1}{L}\abs{\hat{Y}}^2 - \frac{2}{L^2}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) -  \frac{1}{L^3}\abs{ \eta y_k \hat{d}_k^* }^2\\
% &= SS(\{\hat{d}_i, i \in D\}) - R
% \end{align*}
% where 
% \[
% R = \frac{2}{L^2}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) +  \frac{1}{L^3}\abs{ \eta y_k \hat{d}_k^* }^2.
% \]
% We will show that $R$ is positive, from which is will follow that $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$, violating the fact that $\{\hat{d}_i, i \in D\}$ are minimisers of $SS$.

% To see that $R$ is positive write
% \begin{align*}
% R &\geq  \frac{2}{L^2}\abs{\hat{Y} y_k}\Re\left(\eta e^{j(\phi_i - \hat{\theta})}\hat{d}_k^* \right) +  \frac{1}{L^3}\abs{ \eta}^2 \abs{ y_k }^2 \\
% &= \frac{1}{L^2}\abs{y_k} \left( 2\abs{\hat{Y}}\Re\left(\eta e^{j(\phi_i - \hat{\theta})} \hat{d}_k^* \right) + \frac{1}{L}\abs{ \eta}^2 \abs{ y_k }  \right) 
% \end{align*}
% so it is sufficient to show that
% \[
% \Re\left(\eta e^{j(\phi_i - \hat{\theta})} \hat{d}_k^* \right) > - \frac{\abs{ \eta}^2 \abs{ y_k }}{2 L \abs{\hat{Y}}}.
% \]
% Unfinished.  This proof will work, but the problem is that $\abs{y_k}$ can be really small.  The would appear to neccesitate a very large number of buckets.  In some sense, this is no problem, but you have to be able to allocate memory for all of these?  Atleast you need a way of ordering them, up to a certain resolution, without sorting.  I'm not convinced this will ever be possible.
% \end{proof}

>>>>

\begin{theorem}\label{thm:consistency} (Almost sure convergence)
Let $\{w_i\}$ be a sequence of independent and identically distributed, zero mean, circularly symetric complex random variables with $\prob \abs{w_1}$ finite, and let $\{y_i, i \in P \cup D\}$ be observed noisy symbols given by the signal model~\eqref{eq:sigmod}.   Let $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ be the least squares estimator of $a_0 = \rho_0e^{j\theta_0}$ where both $\hat{\rho}$ and $\rho_0$ are positive real numbers.  Put $L = \abs{P \cup D}$ and define real numbers $p$ and $d$ as the limits
\[
\frac{\abs{P}}{L} \rightarrow p \qquad \text{and} \qquad \frac{\abs{D}}{L} \rightarrow d \qquad \text{as $L \rightarrow \infty$.}
\] 
Let $R_i \geq 0$ and $\Phi_i \in [-\pi,\pi)$ be real random variables satisfying
\begin{equation}\label{eq:RiandPhii}
R_ie^{j\Phi_i} = 1 + \frac{1}{\rho_0} w_i e^{-j\theta_0} s_i^*,
\end{equation}
and define the continuous function
\[
G(x) = p h_1(x) + d h_2(x)
\]
where
\[
h_1(x) = \prob R_1 \cos(x + \Phi_1), \;\;\; h_2(x) =  \prob R_1 \cos\sfracpart{ x + \Phi_1}.
\]
If $p > 0$ and if $\abs{G(x)}$ is uniquely maximised at $x = 0$ over the interval $[-\pi,-\pi)$, then:
\begin{enumerate}
\item $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow 0$ almost surely as $L \rightarrow \infty$,
\item $\hat{\rho} \rightarrow \rho_0 G(0)$ almost surely as $L \rightarrow \infty$.
\end{enumerate}
\end{theorem}

@@

\begin{IEEEproof} (Theorem~\ref{thm:normality})
The proof is complicated by the fact that $G_L$ is not differentiable everywhere due to the function $\fracpart{\cdot}$ not being differentiable at half integers.  This precludes the use of ``standard approaches'' to proving normality that are based on the mean value theorem~\cite{vonMises_diff_stats_1947,vanDerVart1971_asymptotic_stats,Pollard_new_ways_clts_1986,Pollard_conv_stat_proc_1984,Pollard_asymp_empi_proc_1989}.  However, we show in Lemma~\ref{lem:diffatlambdaL} that the derivative does exist, and is equal to zero, at the maximiser $\hat{\lambda}_L$, so
\begin{align*}
G_L^\prime(\hat{\lambda}_L) &= 0 \\
&= \frac{1}{L} \sum_{i \in P} R_i \sin(\lambda + \Phi_i) + \frac{1}{L} \sum_{i \in D} R_i \sin\sfracpart{\lambda + \Phi_i}.
\end{align*}
Define the function $Q(\lambda) = \prob G_L^\prime(\lambda) - \prob G_L^\prime(0)$ and write
\begin{align*}
0 &= G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) + Q(\hat{\lambda}_L) \\
&= \sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) + \sqrt{L}Q(\hat{\lambda}_L).
\end{align*}
Lemma~\ref{lem:Qconv} shows that
\[
\sqrt{L} Q(\hat{\lambda}_L) = \sqrt{L} \hat{\lambda}_L\big( ?  + o_P(1) \big)
\]
where $o_P(1)$ denotes a number converging in probablity to zero as $L \rightarrow \infty$.  Lemma~\ref{lem:empiricprocc} shows that
\[
\sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} G_L^\prime(0).
\]
It follows that
\[
0 = o_P(1) + \sqrt{L}G_L^\prime(0) + \sqrt{L} \hat{\lambda}_L \big( ?  + o_P(1) \big)
\]
and rearranging gives,
\[
\sqrt{L} \hat{\lambda}_L = o_P(1) + \frac{\sqrt{L}G_L'(0)}{?  + o_P(1)}.
\]
Lemma~\ref{lem:convdistGLdash} shows that the distribution of $\sqrt{L}G_L'(0)$ converges to the normal with zero mean and variance
\[
p \prob R_1^2 \sin^2(\Phi_1) + d \prob R_1^2 \sin^2\fracpart{\Phi_1}
\]
and it follows that the distribution of $\sqrt{L}\hat{\lambda}_L$ converges to the normal with zero mean and variance
\[
\frac{p \prob R_1^2 \sin^2(\Phi_1) + d \prob R_1^2 \sin^2\fracpart{\Phi_1}}{?}
\]
as required.

@@

\begin{lemma}
\[
\sabs{\sfracpart{\hat{\lambda}_L + \Phi_i}} \leq \frac{1}{2M} - \frac{1}{2LM ?}
\]
for all $i \in D$.
\end{lemma}
\begin{proof}
Recall that $\hat{d}_i = \hat{d}_i(\hat{\theta})$ are the minimsers of $SS$.  The proof now proceeds by contradiction.  Assume that $\sfracpart{\hat{\lambda}_L + \Phi_i} \leq \frac{1}{2M} - \frac{1}{2LM ?}$ for some $k$.  Put $r_i = \hat{d}_i e_k$, then
\begin{align*}
Y(\{ r_i, i \in D\}) = \frac{1}{L}\left( B + \sum_{i \in D} y_i\hat{r}_i^* \right) = \hat{Y} + \frac{1}{L}\eta y_k\hat{d}_k^*.
\end{align*}
where we put $\hat{Y} = Y(\{ \hat{d}_i, i \in D\})$.  Now,
\begin{align*}
SS&(\{r_i, i \in D\}) \\
&= A - \frac{1}{L} \abs{ Y(\{ r_i, i \in D\}) }^2 \\
&= A - \frac{1}{L} \abs{ \hat{Y} + \frac{1}{L} \eta y_k\hat{d}_k^* }^2 \\
&= A - \frac{1}{L}\abs{\hat{Y}}^2 - \frac{2}{L^2}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) -  \frac{1}{L^3}\abs{ \eta y_k \hat{d}_k^* }^2\\
&= SS(\{\hat{d}_i, i \in D\}) - R
\end{align*}
where 
\[
R = \frac{2}{L^2}\Re\left(\eta \hat{Y}^* y_k \hat{d}_k^* \right) +  \frac{1}{L^3}\abs{ \eta y_k \hat{d}_k^* }^2.
\]
We will show that $R$ is positive, from which is will follow that $SS(\{r_i, i \in D\}) < SS(\{\hat{d}_i, i \in D\})$, violating the fact that $\{\hat{d}_i, i \in D\}$ are minimisers of $SS$.

To see that $R$ is positive write
 \begin{align*}
R &\geq  \frac{2}{L^2}\abs{\hat{Y} y_k}\Re\left(\eta e^{j(\phi_i - \hat{\theta})}\hat{d}_k^* \right) +  \frac{1}{L^3}\abs{ \eta}^2 \abs{ y_k }^2 \\
 &= \frac{1}{L^2}\abs{y_k} \left( 2\abs{\hat{Y}}\Re\left(\eta e^{j(\phi_i - \hat{\theta})} \hat{d}_k^* \right) + \frac{1}{L}\abs{ \eta}^2 \abs{ y_k }  \right) 
 \end{align*}
 so it is sufficient to show that
 \[
 \Re\left(\eta e^{j(\phi_i - \hat{\theta})} \hat{d}_k^* \right) > - \frac{\abs{ \eta}^2 \abs{ y_k }}{2 L \abs{\hat{Y}}}.
 \]
 Unfinished, but this proof will work
\end{proof}

\begin{lemma}\label{lem:Qconv}
\[
\sqrt{L} Q(\hat{\lambda}_L) = \sqrt{L} \hat{\lambda}_L\big( ph_1(0) + ?  + o_P(1) \big)
\]
\end{lemma}
\begin{IEEEproof}
Write
\[
Q(\lambda) = \prob G_L^\prime(\lambda) - \prob G_L^\prime(0) = \frac{\abs{P}}{L} k_1(\lambda) + \frac{\abs{D}}{L} k_2(\lambda)
\]
where
\begin{align*}
k_1(\lambda) &= \prob R_1\big( \sin(\lambda + \Phi_1) - \sin(\Phi_1) \big), \;\;\; \text{and} \\
k_2(\lambda) &= \prob R_1\big( \sin\fracpart{\lambda + \Phi_1} - \sin\fracpart{\Phi_1} \big).
\end{align*}
Since
\[
\sin(\lambda + \phi) - \sin(\phi) = \lambda \big( \cos(\phi) + \zeta(\lambda) \big)
\]
where $\zeta(\lambda)$ is a function that goes to zero as $\lambda$ goes to zero, then
\begin{align*}
k_1(\lambda) &= \prob R_1\lambda(\cos(\Phi_1) \; + \; \zeta(\lambda)) \\
&= \lambda \big( \prob R_1\cos(\Phi_1) \; + \; \zeta(\lambda) \prob R_1 \big) \\
&= \lambda \big( h_1(0) \; + \; \zeta(\lambda) K \big),
\end{align*}
where we put $K = \prob R_1$, a finite constant.  Because $\hat{\lambda}_L$ converges almost surely to zero as $L \rightarrow\infty$, then $\zeta(\hat{\lambda}_L)$ converges almost surely (and therefore also in probability) to zero as $L \rightarrow\infty$, i.e. $\zeta(\hat{\lambda}_L) = o_P(1)$.  So, 
\begin{align*}
k_1(\hat{\lambda}_L) &= \hat{\lambda}_L \big( h_1(0) \; + \; \zeta(\hat{\lambda}_L) K \big) \\
&= \hat{\lambda}_L \big( h_1(0)  + o_P(1)\big).
\end{align*}
\end{IEEEproof}

\begin{lemma}\label{lem:empiricprocc} BLERG
\[
\sqrt{L}\big( G_L^\prime(\hat{\lambda}_L) - Q(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} G_L^\prime(0).
\]
\end{lemma}
\begin{IEEEproof}
\end{IEEEproof}

\begin{lemma}\label{lem:convdistGLdash}
The distribution of $\sqrt{L}G_L'(0)$ converges to the normal with zero mean and variance
\[
p \prob R_1^2 \sin^2(\Phi_1) + d \prob R_1^2 \sin^2\fracpart{\Phi_1}.
\]
\end{lemma}
\begin{IEEEproof}
Observe that 
\[
\sqrt{L} G_L^\prime(0) = \frac{1}{\sqrt{L}} \sum_{i \in P} R_i \sin(\Phi_i) + \frac{1}{\sqrt{L}} \sum_{i \in D} R_i \sin\sfracpart{\Phi_i}.
\]
It follows from the standard central limit that the distribution of
\[
\frac{1}{\sqrt{L}} \sum_{i \in P} R_i \sin(\lambda + \Phi_i) = \sqrt{\frac{\abs{P}}{L}} \frac{1}{\sqrt{\abs{P}}} \sum_{i \in P} R_i \sin(\Phi_i) 
\]
converges to the normal with mean $k_1(0) = 0$ and variance 
\[
p \prob R_1^2 \sin^2(\Phi_1).
\]
Similarly, the distribution of 
\[
\frac{1}{\sqrt{L}} \sum_{i \in D} R_i \sin(\lambda + \Phi_i) = \sqrt{\frac{\abs{D}}{L}} \frac{1}{\sqrt{\abs{D}}} \sum_{i \in D} R_i \sin\fracpart{\Phi_i} 
\]
converges to the normal with mean $k_2(0) = 0$ and variance
\[
d \prob R_1^2 \sin^2\fracpart{\Phi_1}.
\]
\end{IEEEproof}

@@
